---
title: "Example Slides From R"
subtitle: "Output to powerpoint!!"
output: 
  powerpoint_presentation:
    slide_level: 3
    reference_doc: ord_science_rmd_template.pptx
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

### R Markdown

This is an R Markdown presentation. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.

### Slide with Bullets

- Bullet 1
- Bullet 2
- Bullet 3

# A section!

### Slide with R Output
::: notes
These are notes that *support*

- a
- little
    - markdown
    
notes on slides with plot output tend to get split up...  Can fix after in ppt.
:::


```{r cars, echo = TRUE}
summary(cars)
```

### Slide with Plot

```{r pressure}
plot(pressure)
```


### Slide with Plot and Bullets in columns 

:::::::::::::: {.columns}
::: {.column}
- These
- Area
- My
- Bullets
:::
::: {.column}
```{r pressure2}
plot(pressure)
```
:::
::::::::::::::
