---
title: "Narragansett Fish Tissue Study"
output:
  html_document:
    df_print: paged
---
This script does exploratory data analysis on concentrations of various analytes in fish tissue, part of EPA TD 4

Client contact: Valerie Bataille

Script by Hayley Brittingham


```{r setup}
require(dplyr)
require(ggplot2)
theme_set(theme_bw())
```

other packages that must be installed (always called by namespace): xlsx, tidyr, scales, ggrepel, gridExtra 

# Read in Data

```{r read in data}
data_2015<-openxlsx::read.xlsx("../data/1.2 Scrubbed Namaus Project Fish Tissue Data 19Mar19.xlsx",
                sheet = "Namaus Project 2015 Data", cols = 1:67)

data_2016<-openxlsx::read.xlsx("../data/1.2 Scrubbed Namaus Project Fish Tissue Data 19Mar19.xlsx",
                           sheet = "Namaus Project 2016 Data", cols = 1:67)

PCB_data_2015<-openxlsx::read.xlsx("../data/1.2 Scrubbed Namaus Project Fish Tissue Data 19Mar19.xlsx",
                          sheet = "2015 Stats", cols = 1:58) 

head(data_2015)
```

* A note about xlsx::read.xslx -- often causes Java errors, especially with newer versions of R

# Look at Variables Available

* Many variables are the same within a species

```{r data tables 2015 species}
data_2015 %>% dplyr::group_by(Species) %>%
  dplyr::summarise(Min_weight=paste(unique(RI_Min_Wt), collapse = ", "),
                   Min_length=paste(unique(RI_Min_Lg), collapse = ", "),
                   Trophic=paste(unique(Trophic), collapse = ", "))
```

* Many more variables are the same within a pond

```{r data tables 2015 ponds}
data_2015 %>% dplyr::group_by(Pond) %>%
  dplyr::summarise(N=length(Trans),
                   Trans=paste(unique(Trans), collapse = ", "),
                   Chloro=paste(unique(Chloro), collapse = ", "),
                   Phos=paste(unique(Phos), collapse = ", "),
                   Nitrogen=paste(unique(Nitrogen), collapse = ", "),
                   pH=paste(unique(pH), collapse = ", "),
                   Oxygen=paste(unique(Oxygen), collapse = ", "),
                   Conduct=paste(unique(Conduct), collapse = ", "),
                   Water=paste(unique(Water), collapse = ", "),
                   Vege=paste(unique(Vege), collapse = ", "),
                   Ecoli=paste(unique(Ecoli), collapse = ", "))
```

# Combine and Organize Data

* A single cohesive dataset will be easier to work with

```{r }
unique(names(data_2015)==names(data_2016))
all_data<-rbind(data_2015, data_2016)
```

* For ease of plotting, want to rename the variables according to the codebook

****QA IMPORTANT****

```{r rename variables}
all_data<-all_data %>% 
  dplyr::mutate(Pond = ifelse(Pond==0, "Schoolhouse Pond", 
                              ifelse(Pond==1, "Deep Pond", 
                                     ifelse(Pond==2, "Upper Schoolhouse Pond", 
                                            ifelse(Pond==3, "Lower Schoolhouse Pond", NA)))), 
                Species=ifelse(Species==1, "Largemouth Bass", 
                               ifelse(Species==2, "Yellow Perch", 
                                      ifelse(Species==3, "Pumpkinseed", 
                                             ifelse(Species==4, "American Eel",
                                                    ifelse(Species==5, "Chain Pickerel",
                                                           ifelse(Species==6, "Brown Bullhead",
                                                                  ifelse(Species==7, "Mussel", NA))))))),
                Sex=ifelse(Sex==1, "Female", ifelse(Sex==2, "Male", ifelse(Sex==8, NA, ifelse(Sex==9, "Pooled Specimen/NA", NA)))),
                Method=ifelse(Method==1, "electrofishing", 
                              ifelse(Method==2, "seine", 
                                     ifelse(Method==3, "rod n reel", 
                                            ifelse(Method==4, "hand harveted", 
                                                   ifelse(Method==9, "Pooled sample/NA",NA))))),
                Guts_Contents=ifelse(Guts_1==0, "Empty Stomach", 
                                     ifelse(Guts_1==1, "Shiner", 
                                            ifelse(Guts_1==2, "Yellow Perch", 
                                                   ifelse(Guts_1==3, "Crayfish", 
                                                          ifelse(Guts_1==4, "Insect", 
                                                                 ifelse(Guts_1==5, "Isopod/Invasive species (woodlice)", 
                                                                        ifelse(Guts_1==6, "Leeches",
                                                                               ifelse(Guts_1==7, "Snapping Turtle", 
                                                                                      ifelse(Guts_1==8, "unidentifiable",
                                                                                             ifelse(Guts_1==9, "pooled sample", 
                                                                                                    ifelse(Guts_1==10, "Shrimp", 
                                                                                                           ifelse(Guts_1==11, "Worms", 
                                                                                                                  ifelse(Guts_1==12, "Killifish", 
                                                                                                                         ifelse(Guts_1==13, "Fish Hook", NA)))))))))))))),
                Lab_Analysis=ifelse(Run==0, "Not analyzed", ifelse(Run %in% c(1:6), "Analyzed", ifelse(Run==7, "Mean", ifelse(Run==8, "Standard Deviation", NA)))),
                Water_level=ifelse(Water==2, "High", ifelse(Water==1, "Medium", ifelse(Water==0, "Low", NA))),
                Vege=ifelse(Vege==2, "Abundant", ifelse(Vege==1, "Normal", ifelse(Vege==0, "Low", NA))))

#organize the PCB data in the same way
PCB_data_2015<-PCB_data_2015 %>% dplyr::filter(Pond %in% c(0,1)) %>%
  dplyr::mutate(Pond = ifelse(Pond==0, "Schoolhouse Pond", 
                              ifelse(Pond==1, "Deep Pond", 
                                     ifelse(Pond==2, "Upper Schoolhouse Pond", 
                                            ifelse(Pond==3, "Lower Schoolhouse Pond", NA)))), 
                Species=ifelse(Species==1, "Largemouth Bass", 
                               ifelse(Species==2, "Yellow Perch", 
                                      ifelse(Species==3, "Pumpkinseed", 
                                             ifelse(Species==4, "American Eel",
                                                    ifelse(Species==5, "Chain Pickerel",
                                                           ifelse(Species==6, "Brown Bullhead",
                                                                  ifelse(Species==7, "Mussel", NA))))))),
                Sex=ifelse(Sex==1, "Female", ifelse(Sex==2, "Male", ifelse(Sex==8, NA, ifelse(Sex==9, "Pooled Specimen/NA", NA)))),
                No=as.numeric(as.character(No))) %>%
  dplyr::select(No, Pond, Species, C_PCBs, NC_PCBs, Year) %>%
  tidyr::gather(key="Analyte", value="Concentration", c("C_PCBs", "NC_PCBs"))

#identify the analytes 
analytes<-c("X.15N", "X.13C", "Se_Hg", "Se_Molar", "Hg_Molar", "Se", "Hg", "As", "Cd", "Co", "Cr", "Cu", "Pb", "Mn", "Ni", "Pt", "Ag", "Sn", "Zn", 
            "EPA", "DHA", "ΣPUFAs")
#rm Se_Hg, molar vars, C13 and N15 for now
analytes<-analytes[-which(analytes %in% c("Se_Hg", "Se_Molar", "Hg_Molar", "X.15N", "X.13C"))]
pcbs<-c("PCB_118_LA", "PCB_126_LA", "PCB_138_LA", "PCB_153_LA", "PCB_180_LA", "ΣPCBs_LA", "PCB_118", "PCB_126", "PCB_138", "PCB_153", "PCB_180", "ΣPCBs")
all_analytes<-c(analytes, pcbs)

#convert the data to a long format for ease of plotting.
all_data_long<-all_data %>% 
  # dplyr::mutate(ΣC_PCBs = as.numeric(as.character(PCB_118)) + as.numeric(as.character(PCB_126)), 
  #                ΣNC_PCBs = as.numeric(as.character(PCB_138)) + 
  #                  as.numeric(as.character(PCB_153)) + as.numeric(as.character(PCB_180))) %>%
  tidyr::gather(key="Analyte", value="Concentration", all_analytes) %>%
  dplyr::mutate(Concentration=as.numeric(as.character(Concentration)),
         EPA_DHA=as.numeric(as.character(EPA_DHA)),
         DHA_EPA=as.numeric(as.character(DHA_EPA))) %>%
  dplyr::filter(No!=0, !is.na(Concentration)) %>%
  dplyr::select(No, Pond, Species, Year, Analyte, Concentration) %>%
  # dplyr::select(-RI_Min_Wt, -RI_Min_Lg, -Lat, -Long, -Guts_1, -Lipid, -Trans, -Chloro,
  #        -Phos, -Nitrogen, -pH, -Oxygen, -Conduct, -Water, -Vege, -Ecoli) %>%
  rbind(., PCB_data_2015) %>%
  dplyr::mutate(Concentration=as.numeric(as.character(Concentration)))

all_data_long$Units<-ifelse(all_data_long$Analyte %in% c("DHA", "EPA", "ΣPUFAs"), "(mg/g)", 
                            ifelse(all_data_long$Analyte %in% analytes, "(μg/g)",
                                   ifelse(all_data_long$Analyte %in% c(pcbs, "NC_PCBs", "C_PCBs"), "(ng/g)", NA)))
all_data_long$Analyte_lab<- paste(all_data_long$Analyte, all_data_long$Units)

head(all_data_long)
```

## Initial Plots

```{r bad plots 1}
ggplot(all_data_long[which(all_data_long$Analyte %in% analytes),], 
       aes(x=Analyte_lab, y=Concentration)) +
  geom_boxplot() + geom_point(shape=21) +
  labs(y="Concentration (dry weight)") + 
  scale_y_continuous(trans="log10") + 
  annotation_logticks(sides="l") +
  theme(axis.text.x = element_text(angle=60, vjust=0.7)) 
```

```{r bad plots 2, fig.height=8}
ggplot(all_data_long[which(all_data_long$Analyte %in% analytes),], 
       aes(x=Pond, y=Concentration)) +
  geom_boxplot() + geom_point(aes(color=Pond)) + 
  facet_wrap(~Analyte_lab, scales="free_y") + labs(y="Concentration (dry weight)") + 
  ggtitle("All Species: non-PCBs") +
  theme(axis.text.x = element_text(angle=90), plot.title=element_text(hjust=0.5),
        legend.position = "none")
```

## Revised Plots

* Bar charts

```{r bar Hg, fig.height=8}
#In this case, more interested in species comparisons than pond comparisons

#summarize data for a bar graph #first by species
bar_dat<-all_data_long %>% dplyr::group_by(Analyte_lab, Analyte, Pond, Species) %>% 
  summarise(mean_conc=mean(Concentration, na.rm=T), 
            n=length(which(!is.na(Concentration))),
            SE=sd(Concentration, na.rm = T)/sqrt(n))

#bar graphs concentration vs species, faceted by analyte
ggplot(bar_dat[which(bar_dat$Analyte %in% analytes & bar_dat$Pond=="Deep Pond"),]) +
  geom_bar(aes(x=Species, y=mean_conc, fill=Species), stat="identity") + 
  geom_errorbar(aes(x=Species, ymin=mean_conc-SE, ymax=mean_conc + SE)) +
  facet_wrap(~Analyte_lab, scales="free_y") + labs(y="Concentration (dry weight)") + 
  ggtitle("All Species: non-PCBs") +
  theme(axis.text.x = element_text(angle=90), plot.title=element_text(hjust=0.5),
        legend.position = "none")

#we want the legend for the species to be the same for all plots. 
color.vec<-scales::hue_pal()(length(unique(bar_dat$Species)))
names(color.vec)=unique(factor(bar_dat$Species))
color.vec
```

```{r bar graph Hg}
#bar graph by species, Hg only.
ggplot(bar_dat[which(bar_dat$Analyte =="Hg"),]) +
  geom_bar(aes(x=Species, y=mean_conc, fill=Species), stat="identity", show.legend = F) + 
  geom_errorbar(aes(x=Species, ymin=mean_conc-SE, ymax=mean_conc + SE)) +
  facet_wrap(~Pond) + labs(y="Hg Concentration (dry weight, ug/g)") + 
  ggtitle("Mercury Concentration by Pond and Species") +
  geom_hline(aes(yintercept=0.049, linetype="Subsistence")) +
  geom_hline(aes(yintercept=0.4, linetype="Recreational")) +
  theme(axis.text.x = element_text(angle=60, vjust=.5, size=11), 
        axis.title = element_text(size=12), axis.text.y = element_text(size=11),
        strip.text = element_text(size=13),
        plot.title=element_text(hjust=0.5)) +
  scale_fill_manual(values = color.vec) + labs(linetype="")

```

* Same idea as a pie chart
  * Emphasizes relative nature of research goals
  * Looking at contamination as a whole
  
```{r }
  #calculate percentages for a pie chart
#schoolhouse pond
pie_dat_SH<-all_data_long %>% dplyr::group_by(Analyte_lab, Analyte, Pond, Species) %>% 
  dplyr::filter(Pond=="Schoolhouse Pond", Analyte=="Hg") %>%
  dplyr::summarise(sum_conc=sum(Concentration), Nfish=length(Concentration)) %>%
  dplyr::mutate(perc_conc=sum_conc/sum(sum_conc), perc_conc_label=paste0(round(perc_conc,2)*100, "%")) %>%
  dplyr::arrange(-perc_conc) %>%
  dplyr::mutate(Species=factor(Species, levels = unique(Species))) %>%
  dplyr::arrange(perc_conc) %>%            
  dplyr::mutate(cum_sum=cumsum(perc_conc))

pie_dat_SH$mid_sum<-pie_dat_SH$cum_sum/2
for (i in 2:nrow(pie_dat_SH)) {
  #just played with these numbers such that the text was always readable and towards the middle of the slice
  adj<-ifelse(i==2, 0.03, ifelse(i==3, -.01, 0)) 
  pie_dat_SH$mid_sum[i]<-(pie_dat_SH$cum_sum[i-1] + pie_dat_SH$cum_sum[i]) /2 + adj
}

head(pie_dat_SH)
pie_SH<-ggplot(pie_dat_SH) +
  geom_bar(aes(x=0, y=perc_conc, fill=Species, group=Species), stat="identity") + 
  # ggrepel::geom_text_repel(aes(x=0, y=mid_sum, label=perc_conc_label)) +
  ggrepel::geom_text_repel(data=pie_dat_SH[which(pie_dat_SH$Species!="Brown Bullhead"),],
                           aes(x=0, y=mid_sum, label=perc_conc_label)) +
  ggrepel::geom_text_repel(data=pie_dat_SH[which(pie_dat_SH$Species=="Brown Bullhead"),],
                           aes(x=0.2, y=mid_sum, label=perc_conc_label), nudge_x = 0.8) +
  labs(y="", x="") + 
  ggtitle("Schoolhouse Pond") +
  theme(axis.ticks = element_blank(), plot.title=element_text(hjust=0.5), axis.text = element_blank()) +
  coord_polar("y", start=0) +
  scale_fill_manual(values=color.vec)
pie_SH

#same for Deep pond
pie_dat_DP<-all_data_long %>% dplyr::group_by(Analyte_lab, Analyte, Pond, Species) %>% 
  dplyr::filter(Pond=="Deep Pond", Analyte=="Hg") %>%
  dplyr::summarise(sum_conc=sum(Concentration), Nfish=length(Concentration)) %>%
  dplyr::mutate(perc_conc=sum_conc/sum(sum_conc), perc_conc_label=paste0(round(perc_conc,2)*100, "%")) %>%
  dplyr::arrange(-perc_conc) %>%
  dplyr::mutate(Species=factor(Species, levels = unique(Species))) %>%
  dplyr::arrange(perc_conc) %>%            
  dplyr::mutate(cum_sum=cumsum(perc_conc))

pie_dat_DP$mid_sum<-pie_dat_DP$cum_sum/2
for (i in 2:nrow(pie_dat_DP)) {
  #just played with these numbers such that the text was always readable and towards the middle of the slice
  adj<-ifelse(i==2, 0.025, ifelse(i==3, -.01, 0.01))
  pie_dat_DP$mid_sum[i]<-(pie_dat_DP$cum_sum[i-1] + pie_dat_DP$cum_sum[i]) /2 + adj
}

pie_DP<-ggplot(pie_dat_DP[which(pie_dat_DP$Analyte =="Hg" & pie_dat_DP$Pond=="Deep Pond"),]) +
  geom_bar(aes(x=0, y=perc_conc, fill=Species, group=Species), stat="identity") + 
  ggrepel::geom_text_repel(data=pie_dat_DP[which(pie_dat_DP$Species!="Brown Bullhead"),],
                           aes(x=0, y=mid_sum, label=perc_conc_label)) +
  ggrepel::geom_text_repel(data=pie_dat_DP[which(pie_dat_DP$Species=="Brown Bullhead"),],
                           aes(x=0.2, y=mid_sum, label=perc_conc_label), nudge_x = 0.8) +
  labs(y="", x="") + 
  ggtitle("Deep Pond") +
  theme(axis.ticks = element_blank(), plot.title=element_text(hjust=0.5), axis.text = element_blank()) +
  coord_polar("y", start=0) +
  scale_fill_manual(values=color.vec)
pie_DP
```

## Plots for PCBs

```{r PCBs, fig.width=10}
bar_dat_sum<-all_data_long %>% dplyr::group_by(Analyte_lab, Analyte, Pond, Units) %>% 
  dplyr::summarise(mean_conc=mean(Concentration, na.rm = TRUE), SE=sd(Concentration, na.rm = TRUE)/sqrt(length(!is.na(Concentration)))) %>%
  dplyr::mutate(Analyte2=ifelse(Analyte %in% pcbs, gsub("_LA", "", gsub("PCB_", "", Analyte)),
                               ifelse(Analyte=="Hg", "Hg", as.character(Analyte))),
                An_cat=ifelse(Units=="(μg/g)", "Other Analytes (μg/g)", 
                              ifelse(grepl("LA", Analyte), "Lipid Adjusted PCBs (ng/g)", 
                                           "PCBs (ng/g)")),
                An_cat=factor(An_cat, levels = c("Lipid Adjusted PCBs (ng/g)", "PCBs (ng/g)", "Other Analytes (μg/g)")),
                An_cat2 = ifelse(An_cat=="PCBs (ng/g)", "PCBs - 2015 only (ng/g)", as.character(An_cat)),
                An_cat2=factor(An_cat2, levels = c("Lipid Adjusted PCBs (ng/g)", "PCBs - 2015 only (ng/g)", "Other Analytes (μg/g)")),
                SL=ifelse(Analyte2=="Hg", 0.049, 
                          ifelse(Analyte2=="Se", 2.457,
                                 ifelse(Analyte2=="NC_PCBs", 2.45, 
                                        ifelse(Analyte2=="C_PCBs", 9.8, NA)))),
                Analyte_lab2=ifelse(Analyte2=="Hg", "Mercury", 
                                   ifelse(Analyte2=="Se", "Selenium", 
                                          ifelse(Analyte2=="C_PCBs", "Sum Carcinogenic PCBs", 
                                                 ifelse(Analyte2=="NC_PCBs", "Sum Non-Carcinogenic PCBs", as.character(Analyte2)))))) %>%
  dplyr::filter(Analyte %in% c("Hg","Se","NC_PCBs","C_PCBs", pcbs))

'%!in%' <- function(x,y)!('%in%'(x,y))

ggplot(bar_dat_sum[which(bar_dat_sum$Analyte2 %in% c("Hg", "Se", "C_PCBs", "NC_PCBs") &
                           bar_dat_sum$An_cat2=="PCBs - 2015 only (ng/g)"),]) +
  geom_bar(aes(x=Analyte_lab2, y=mean_conc, fill=Analyte_lab2), stat="identity") + 
  geom_errorbar(aes(x=Analyte_lab2, ymin=mean_conc-SE, ymax=mean_conc + SE)) +
  facet_wrap(~Pond) +labs(y="Mean Concentration (dry weight, ng/g)", x="Analyte") + 
  theme(plot.title=element_text(hjust=0.5), axis.text = element_text(size=12), 
        axis.title = element_text(size=13), strip.text = element_text(size=13),
        legend.position = "none") +
  geom_errorbar(aes(x=Analyte_lab2, ymin=SL, ymax=SL,color="Screening Level")) +
  scale_color_manual(values="red") + scale_fill_brewer(palette = "Accent")

```

```{r by species}
ggplot(bar_dat[which(bar_dat$Analyte == "C_PCBs"),]) +
  geom_bar(aes(x=Species, y=mean_conc, fill=Species), stat="identity") + 
  geom_errorbar(aes(x=Species, ymin=mean_conc-SE, ymax=mean_conc + SE)) +
  facet_wrap(~Pond) + labs(y="Concentration (dry weight, ng/g)") + 
  ggtitle("Carcinogenic PCBs Concentration by Pond and Species") +
  geom_hline(aes(yintercept=9.8, linetype="Subsistence")) +
  # geom_hline(aes(yintercept=20, linetype="Recreational")) +
  theme(axis.text.x = element_text(angle=60, vjust=.5), plot.title=element_text(hjust=0.5),
        legend.position = "none") +
  scale_fill_manual(values = color.vec)

```